Toward nonlinear multigrid \\ for nonlinear variational inequalities

### 2023-02-06 talk/slides.tex
start multigrid material; other structure

\bigskip
\only<1>{
\begin{itemize}
\item \emph{problem.} on a domain $\Omega \subset \RR^2$, find the displacement $u(x)$ of a membrane, with fixed value $u = g$ on $\partial \Omega$, above an \emph{obstacle} $\psi(x)$, which minimizes the elastic energy
    $$J(u) = \int_\Omega \frac{1}{2} |\grad u|^2 - f\, u$$
\item shown above: $\Omega=(-2,2)^2$, $\psi(x)$ a hemisphere
\end{itemize}
}
\only<2>{
\begin{itemize}
\item i.e.~constrained optimization over an \emph{admissible set}
	$$\mathcal{K} = \left\{v \in H^1(\Omega) \,:\, v\big|_{\partial \Omega} = g \text{ and } v \ge \psi\right\}$$
\item informally, $J'(u)$ points directly into $\mathcal{K}$ (\emph{variational inequality}; VI):
    $$\ip{J'(u)}{v-u} = \int_\Omega \grad u\cdot \grad (v-u) - f (v-u) \ge 0 \quad \text{for all } v \in \mathcal{K}$$
\end{itemize}
}
\only<3>{
\begin{itemize}
\item the solution defines \emph{active} $A_u = \{u = \psi\}$ and \emph{inactive} $R_u = \{u> \psi\}$ subsets of $\Omega$, and a \emph{free boundary} $\Gamma_u=\partial R_u \cap \Omega$
\item a naive strong form poses the problem in terms of its solution:
{\small
\begin{align*}
-\grad^2 u &= f \text{ on $R_u$} \\
u &= \psi \text{ on $A_u$}
\end{align*}
}
\item by contrast, the weak form VI is meaningful
\end{itemize}
}
\only<4>{
\begin{itemize}
\item a \emph{complementarity problem}, essentially KKT conditions, is a meaningful strong form:
{\small
\begin{align*}
u - \psi &\ge 0 \\
-\grad^2 u - f &\ge 0 \\
(u - \psi)(-\grad^2 u - f) &= 0
\end{align*}
}
\end{itemize}
}
\end{frame}

@@

\begin{itemize}
\item let $\mathcal{K}$ be a closed and convex subset of a Banach space $\mathcal{V}$
\item suppose $F:\mathcal{K} \to \mathcal{V}'$ is a continuous, generally nonlinear \emph{operator}
    \begin{itemize}
    \item[$\circ$] $F$ may be defined only on $\mathcal{K}$
    \item[$\circ$] $F$ may \emph{not} be the derivative of a $J(u)$
    \end{itemize}
\item the general problem {\color{FireBrick} VI($F$,$\mathcal{K}$)} is
	$${\color{FireBrick} \ip{F(u)}{v-u} \ge 0 \quad \text{ for all } v \in \mathcal{K}}$$
\end{itemize}
\end{frame}

@@

\begin{center}
\begin{tabular}{l|l}
\begin{minipage}[t][16mm][t]{0.4\textwidth}
unconstrained optimization:
$$\min_{u\in\mathcal{V}} J(u)$$
\end{minipage}
&
\begin{minipage}[t][16mm][t]{0.4\textwidth}
constrained optimization:
$$\min_{u\in\mathcal{K}} J(u)$$
\end{minipage}
\\ \hline
\begin{minipage}[t][16mm][t]{0.4\textwidth}
solve equation for $u \in \mathcal{V}$: {\LARGE \strut}

@@

\vspace{-2mm}
solve {\color{FireBrick} VI($F$,$\mathcal{K}$)} for $u \in \mathcal{K}$:
$${\color{FireBrick} \ip{F(u)}{v-u} \ge 0 \quad \forall v \in \mathcal{K}}$$
\end{minipage}
\end{tabular}
\end{center}
\end{frame}

\begin{frame}{most familiar case}

\begin{itemize}
\item we even teach VIs when we introduce calculus

\includegraphics[width=\textwidth]{images/calcone.png}
\end{itemize}
\end{frame}

@@

>>>>

\bigskip
\only<1>{
\begin{itemize}
\item \emph{problem.} on a domain $\Omega \subset \RR^2$, find the displacement $u(x)$ of a membrane, with fixed value $u = g$ on $\partial \Omega$, above an \emph{obstacle} $\psi(x)$, which minimizes the elastic energy
    $$J(u) = \int_\Omega \frac{1}{2} |\grad u|^2 - f\, u$$
\item shown above: $\Omega=(-2,2)^2$, $\psi(x)$ a hemisphere
\end{itemize}

\phantom{x}
}
\only<2>{
\begin{itemize}
\item i.e.~constrained optimization over an \emph{admissible set}
	$$\mathcal{K} = \left\{v \in H^1(\Omega) \,:\, v\big|_{\partial \Omega} = g \text{ and } v \ge \psi\right\}$$
\item informally, $J'(u)$ points directly into $\mathcal{K}$ ({\color{FireBrick} \emph{variational inequality}; VI}):
    $$\ip{J'(u)}{v-u} = \int_\Omega \grad u\cdot \grad (v-u) - f (v-u) \ge 0 \quad \text{for all } v \in \mathcal{K}$$
\end{itemize}
}
\only<3>{
\begin{itemize}
\item the solution defines \emph{active} $A_u = \{u = \psi\}$ and \emph{inactive} $R_u = \{u> \psi\}$ subsets of $\Omega$, and a \emph{free boundary} $\Gamma_u=\partial R_u \cap \Omega$
\item a naive strong form poses the problem in terms of its solution:
{\small
\begin{align*}
-\grad^2 u &= f \text{ on $R_u$} \\
u &= \psi \text{ on $A_u$}
\end{align*}
}
\item by contrast, the weak form VI is meaningful

\phantom{x}
\end{itemize}
}
\only<4>{
\begin{itemize}
\item a \emph{complementarity problem} (CP) is a meaningful strong form:
{\small
\begin{align*}
u - \psi &\ge 0 \\
-\grad^2 u - f &\ge 0 \\
(u - \psi)(-\grad^2 u - f) &= 0
\end{align*}
}
\item for optimization problems: CP $=$ KKT conditions

\phantom{x}

\phantom{x}
\end{itemize}
}
\end{frame}

@@

\begin{itemize}
\item let $\mathcal{K}$ be a closed and convex subset of a Banach space $\mathcal{V}$
\item suppose $F:\mathcal{K} \to \mathcal{V}'$ is a continuous, generally nonlinear operator
    \begin{itemize}
    \item[$\circ$] $F$ may be defined only on $\mathcal{K}$
    \item[$\circ$] $F$ may \emph{not} be the derivative of a $J(u)$
    \end{itemize}
\item the general problem {\color{FireBrick} VI($F$,$\mathcal{K}$)} is
	$${\color{FireBrick} \ip{F(u)}{v-u} \ge 0 \quad \text{ for all } v \in \mathcal{K}}$$
\end{itemize}
\end{frame}

@@

\begin{center}
\begin{tabular}{l|l}
\begin{minipage}[t][16mm][t]{0.4\textwidth}
unconstrained optimization:
$$\min_{u\in\mathcal{V}} J(u)$$
\end{minipage}
&
\begin{minipage}[t][16mm][t]{0.4\textwidth}
constrained optimization:
$$\min_{u\in\mathcal{K}} J(u)$$
\end{minipage}
\\ \hline
\begin{minipage}[t][16mm][t]{0.4\textwidth}
equation for $u \in \mathcal{V}$: {\LARGE \strut}

@@

\vspace{-2mm}
{\color{FireBrick} VI($F$,$\mathcal{K}$)} for $u \in \mathcal{K}$:
$${\color{FireBrick} \ip{F(u)}{v-u} \ge 0 \quad \forall v \in \mathcal{K}}$$
\end{minipage}
\end{tabular}
\end{center}
\end{frame}

\begin{frame}{applications of VIs}

\begin{itemize}
\item elastic contact, Signorini problems (e.g.~Kikuchi \& Oden 1988)
\item viscous contact problems (de Diego et al.~2022)
\item pricing of American options in the Black-Scholes model
\item the geometry of glaciers \dots \emph{more soon}

\bigskip
\item first-semester calculus
\end{itemize}

\begin{center}
\includegraphics[width=0.9\textwidth]{images/calcone.png}
\end{center}
\end{frame}

\AtBeginSection[]
{
  \begin{frame}{Outline}
    \tableofcontents[currentsection]
  \end{frame}
}

@@

\begin{frame}{a nonlinear 2-grid scheme}

\begin{center}
\includegraphics[width=0.3\textwidth]{images/two-grids.png}
\end{center}

\bigskip
\only<1>{
\begin{itemize}
\item suppose we have a nonlinear elliptic PDE problem:
	$$F(u) = 0$$
\end{itemize}

\phantom{x}
}
\only<2>{
\begin{itemize}
\item x
\end{itemize}
}
\end{frame}

\begin{frame}{nonlinear multigrid}

\mbox{
\includegraphics[height=0.10\textheight]{images/mg-grids.png} \qquad \includegraphics[height=0.15\textheight]{images/mg-vcycle.png} \quad
\includegraphics[height=0.15\textheight]{images/mg-fcycle.png}
}

\bigskip
\only<1>{
\begin{itemize}
\item suppose we have a nonlinear elliptic PDE problem:
	$$F(u) = 0$$
\end{itemize}

\phantom{x}
}
\only<2>{
\begin{itemize}
\item x
\end{itemize}
}
\end{frame}

\begin{frame}{nonlinear elliptic PDE problem}
